\section{I wanna learn more!}

Cool, here's a few pointers.
I am also working on an interactive learning platform that teaches mathematical techniques for linguistics, and \textbf{\textcolor{Red3}{I am looking for volunteers}} to test it and provide feedback.
If you are interested, or if you have any questions on today's material, shoot me a line: \href{mailto:mail@thomasgraf.net}{mail@thomasgraf.net}

\begin{itemize}
    \item \textbf{Automata Theory}
        \begin{itemize}
            \item \emph{Introduction to the Theory of Computation}\\
                undergraduate course at Stony Brook, Computer Science
            \item Michael Sipser. 2012. 3rd edition.\\
                \emph{Introduction to the Theory of Computation}. (yep, that's the same name)

                \smallskip
                \emph{Pro tip:} The 2nd edition is almost the same and used copies are much cheaper.
                A decent scan can easily be found on the first Google page, I'm not sure about its legality.
        \end{itemize}
    \item \textbf{Computational Linguistics}
        \begin{itemize}
            \item \emph{North American Computational Linguistics Olympiad}\\
                we have training sessions at Stony Brook
            \item \emph{Language and Technology}\\
                undergraduate course at Stony Brook, Linguistics
            \item \emph{Computational Linguistics}\\
                undergraduate course at Stony Brook, Linguistics
            \item \emph{Natural Language Processing}\\
                undergraduate course at Stony Brook, Computer Science
            \item Markus Dickinson, Chris Brew, and Detmar Meurers. 2012.\\
                \emph{Language and Computers}.

                \smallskip
                Used copies are available for 10 bucks on Amazon.
        \end{itemize}
    \item \textbf{Language from a Formal and Cognitive Perspective}
        \begin{itemize}
            \item Check out my website, teaching materials can be found under \emph{Students}
            \item Steven Pinker. \emph{The Language Instinct: How the Mind Creates Language}.
        \end{itemize}
    \item \textbf{The Human Mind as a Computer}
        \begin{itemize}
            \item Douglas Hofstadter. \emph{Gödel, Escher, Bach: An Eternal Golden Braid}.
        \end{itemize}
\end{itemize}
